---
title: "Housing and Net-Zero"
author: "Professor Adam Dennett - @adam_dennett"
institute: Bartlett Centre for Advanced Spatial Analysis, University College London
date: today
format: 
  revealjs:
    theme: solarized
    chalkboard: true
    smaller: true
    logo: images/University_College_London_logo.svg
    transition: slide
    transition-speed: fast
    width: 1600
    height: 900
    footer: | 
      <https://github.com/adamdennett/EPC_Data_Analysis> 
date-format: long
editor: visual
slide-number: true
---

## UK CO^2^ Emissions

```{r}
library(tidyverse)
library(readxl)
library(ggplot2)

#download UK Greenhouse Gas Data from here: https://www.gov.uk/government/statistics/final-uk-greenhouse-gas-emissions-national-statistics-1990-to-2022

#data below taken from here: https://assets.publishing.service.gov.uk/media/65c0d17663a23d0013c821ea/final-greenhouse-gas-emissions-tables-2022.xlsx

GreenhouseGasProp <- read_excel("data/UK_CO2.xlsx", sheet=3)
GreenhouseGasTotal <- read_excel("data/UK_CO2.xlsx", sheet=4)

GreenhouseGasProp_Long <- pivot_longer(GreenhouseGasProp,
                                       cols = 2:34, 
                                       names_to = "year",
                                       values_to = "percent")

#GreenhouseGasProp_Long$year <- as.numeric(GreenhouseGasProp_Long$year)

#ggplot(GreenhouseGasProp_Long, aes(x=year, y=percent,fill = Sector)) + geom_area()

GreenhouseGasTotal_Long <- pivot_longer(GreenhouseGasTotal,
                                       cols = 2:34, 
                                       names_to = "year",
                                       values_to = "total")

GreenhouseGasTotal_Long$year <- as.numeric(GreenhouseGasTotal_Long$year)

ggplot(GreenhouseGasTotal_Long, aes(x=year, y=total,fill = Sector)) + geom_area() + labs(x = "Year", y = "Million tons carbon dioxide equivalent (MtCO2e)", title = "UK Greenhouse Gas Emissions by Source Category")

```

Heating (and , increasingly, cooling) Residential Buildings currently accounts for about 13.7% of the UK's total greenhouse gas emissions.

While Greenhouse Gas emissions have reduced since 1990 (largely thanks to renewables becoming more important in generating electricity) housing is now responsible for a greater proportion of our total emissions than ever before - producing around 56 million tons of CO^2^ per year

## UK Housing and CO^2^ Emissions

```{r}
#get some data on how much better Europe is than the UK- https://ec.europa.eu/eurostat/databrowser/view/ENV_AC_AINAH_R2__custom_10056783/default/table?lang=en

library(janitor)

EU_Co2Housing <- read_csv("data/EU_CO2_Housing.csv")

EU_Co2Housing22 <- EU_Co2Housing[,c(1,14)]
colnames(EU_Co2Housing22) <- c("Country", "Year_2022")

EU_Co2Housing_Long <- pivot_longer(EU_Co2Housing,
                                       cols = 2:14, 
                                       names_to = "year",
                                       values_to = "total")

# ggplot(EU_Co2Housing_Long, aes(x=year, y=total, group = Country, colour = Country)) + geom_line() + labs(x = "Year", y = "Million tons carbon dioxide equivalent (MtCO2e)", title = "EU Residential Greenhouse Gas Emissions")

options(scipen=10000)
ggplot(EU_Co2Housing22) + geom_col(aes(y= reorder(Country, Year_2022), x=Year_2022, fill = Year_2022), show.legend = FALSE) + labs(y = "Country", x = "Tons carbon dioxide equivalent (MtCO2e)", title = "EU Residential Heating/Cooling Emissions") + scale_fill_viridis_c(option = "magma")

```

Housing in the UK is some of the worst, overall, in Europe for emitting greenhouse gases. 

Source: Eurostat + UK Government

## UK Housing and CO^2^ Emissions

```{r}
EU_Co2Housing_pp <- read_csv("data/EU_CO2_Housing_per_person.csv")

options(scipen=10000)
ggplot(EU_Co2Housing_pp) + geom_col(aes(y= reorder(Country, Res_CO2_per_Person), x=Res_CO2_per_Person, fill = Res_CO2_per_Person), show.legend = FALSE) + labs(y = "Country", x = "Tons carbon dioxide equivalent (MtCO2e) per person", title = "EU Residential Heating/Cooling Emissions Per Person") + scale_fill_viridis_c(option = "magma")

```

Even after adjusting for population, UK housing stock is responsible for emitting far more CO2 than housing in very cold countries like Iceland, Sweden, Norway and Finland

Source: Eurostat + UK Government

## Leaky Homes

In 2021, the National Housing Federation reported that a combination of gas central heating and poor insulation meant that as heat easily leaks from many of the UK's homes, it requires even more gas to heat them.

The average family or household in England is currently producing more CO^2^ every year by just living in their home than they are by driving. 

Three in every five (60%) homeowners don’t think their home energy use has much of an impact on carbon emissions. More than one in four homeowners (28%) have no plans to make “eco-upgrades” to their homes in the next ten years.

https://www.housing.org.uk/news-and-blogs/news/englands-leaky-homes-greater-threat-to-climate-than-cars/

##